\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-04 paper.tex
Edits from Yan and Gregor

We show that a synthesis of these views can be achieved by thinking
in terms of genomes and transmission
intervals.
% Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
% based on exhaustively detailing the ``events'' that occur
% in the history of a sample,
%are insufficient to describe the outputs of current methods.
Moreover, we argue that the underlying assumption
that all events can be known and precisely estimated
is fundamentally unsuited to the realities
of modern, population-scale datasets.
We propose an alternative mathematical formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

The breakthroughs in scale achieved by recent
methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history.
% This is a digression from the maing point of the paragraph, so
% making it a parenthetical aside.
(Note that it is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.)
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

We address this problem by providing a simple formal definition of an
ARG data structure, based on recording the intervals of
genetic inheritance between specific genomes.
We call this the ``genome ARG'', or gARG encoding. We
contrast this with the classical formal definition of an ARG, based on
recording common ancestor and recombination events, which we refer to as the
``event ARG'' or eARG encoding. We show that the new gARG encoding is a
substantial generalisation of the classical eARG approach,  providing much more
flexibility in how genetic inheritance can be represented, and encompasses the
outputs of modern methods. We show that the gARG approach can represent many
different types of approximation, in particular allowing us to systematically
describe uncertainty about the temporal ordering of multiple recombinations. It
is important to note that throughout we are interested in the details of these
competing mathematical formulations and their practical consequences.
% True ARGs, being perfectly resolved, can ultimately be described in
% either formalism

@@

% % I feel like sophisticated readers would at this point be saying,
% % ah, yeah, that's basically what an ARG is just in different words;
% % what's the point of this paper?
The genome ARG framework defined here is
in many ways simply a clarification of existing treatments
\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
adding concrete details to describe the
differential inheritance of genetic material between genomes.
% % This is definitely needed for the Nicks, who are always thinking
% % about the deeper processes and are quite happy being loosey-goosey
% % with representation details
It is important to note that here, and throughout,
we are not questioning the form of the
actual ancestral processes that occur in nature, but rather how we
\emph{represent} the outcomes of such processes in a practical manner.
% Again, need to remind the reader that there's an actual point
% to all this getting stuck into the details.
These practical details,
as demonstrated in later sections, have important
consequences not only for how methods exchange information about
simulated and inferred ARGs, but more fundamentally in
how we set our goals for inference and evaluate the success of results.

\section{Event ARGs}
\label{sec-eARG}
In this section we define a common (and classical) view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and used as an output format by multiple ARG inferences tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this encoding we have two types of internal node in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this encoding as the ``event ARG'' or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

% eARGs are representationally for limited interchange
While this approach of annotating recombination nodes with a
breakpoint in an eARG is a concise and elegant way of describing realisations
of the coalescent, it is inherently limited when
implemented literally. The eARG encoding explicitly models only
two different types of event and thus anything that is not a single crossover
recombination or common ancestor event, must be incorporated
either in a roundabout way using these
events, or by adding new types of event to the encoding. For example, gene
conversion could be accommodated either by stipulating a third type of event
(annotated by two breakpoints and corresponding traversal conventions for
recovering the local trees) or by two recombination nodes joined by a
zero-length edge. The gARG encoding described in the previous
section offers a simpler and more direct solution.

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG (here, and throughout, we use the general term ``ARG'' when
the details of the specific encoding are not important).
It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
For example, in Fig.~\ref{fig-arg-in-pedigree} we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of their ancestral
material propagating through the graph
backwards in time. In the region $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find a common ancestor in \noderef{f}. Thus, in this region,
the total number of genome segments that are ancestral to the
sample is reduced from four to three. Fig.~\ref{fig-arg-in-pedigree}A
illustrates this by (shaded) ancestral material being present
in only three nodes (\noderef{f}, \noderef{g}, and \noderef{h}) in this region,
while node \noderef{e} is blank
as it carries \emph{non-ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

@@

Efficiently propagating and resolving ancestral material for
a sample through a pre-existing graph is a well-studied problem,
and central to recent advances in individual-based forward-time
simulations~\citep{kelleher2018efficient,haller2018tree}.
In contrast to the usual ``retrospective'' view of ARGs
discussed so far, these methods record an ARG forwards in
time in a ``prospective'' manner. Genetic inheritance relationships
and mutations are recorded exhaustively, generation-by-generation,
leading to a rapid build-up of information, much of which
will not be relevant to the genetic ancestry of the current population.
This redundancy is periodically removed using the ``simplify''
algorithm~\citep{kelleher2018efficient}, which propagates and
resolves ancestral material.
Efficient simplification is the key enabling factor for
this prospective-ARG based approach to forward-time simulation,
which can be orders of magnitude faster than standard
sequence-based methods
(see Section~\ref{sec-ARG-simplification} for
other applications of ARG simplification).
% Define "sample-resolved" here for later use
We refer to a gARG that has been simplified with respect to a set of
samples, such that the inheritance annotations on its edges contain
no non-ancestral material,  as sample-resolved.

@@

Any eARG can be converted to a sample-resolved gARG
via a two-step process illustrated in Fig.~\ref{fig-ancestry-resolution}.
The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A),
duplicate its graph topology, and then add inheritance annotations
to each of the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B) as follows.
If a given node is a common ancestor event, we annotate the single
outbound edge with the interval $[0,L)$, for a genome of length $L$. If the
node is a recombination event with a breakpoint $x$, we annotate the two
outbound edges respectively with the intervals $[0, x)$ and $[x, L)$. These
inheritance interval annotations are clearly in one-to-one correspondence with
the information in the input eARG. They are also analogous to the
inheritance intervals we get on the edges in a prospective gARG
produced by a forward-time simulation, which are concerned with recording
the direct genetic relationship between a parent and child genome and are not
necessarily minimal in terms of the resolved ancestral material of a sample.
Thus, the final step is to use the ``simplify'' algorithm to perform the required
sample resolution (Fig.~\ref{fig-ancestry-resolution}C).

@@

3 PARAS. - enough outline to let readewr know what's in the appendices.

There are lots of new inference techniques: they all produce slightly different
structures, all of which can be though of a types of ARGs...

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate. See the
text for details of these methods.
Edge colours indicate time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Bottom row graphics show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines denote breakpoints between
local trees.
}
\end{figure}

These give different levels of precision about recombination (see new Appendix X)

You can see this as different levels of detail on the same underlying truth. Here's a figure:

@@

The timing, positions, and even the number of recombination events is generally
not possible to infer precisely from genome sequencing data. Under
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation and recombination
rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
position can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region.
% For general
% ARGs, for instance those representing pedigrees, or genealogies of organisms with
% low mutation rates and complex patterns of recombination, such inference can be
% even more challenging.
The fact that the eARG encoding \emph{requires}
precise information about recombination is therefore a fundamental limitation.

\section{Discussion}
\label{sec-discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in inference methods, their evaluation, and their application to biological questions.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix~\ref{sec-arg-history}, the term
``ancestral recombination graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definition
within the context of stochastic
processes and instead use the term as a more general representation of any
recombinant genetic ancestry seems useful; we have
tried to clarify and systematise it here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG (gARG) encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG) encoding, which is more limited in what can be described.
Importantly, gARGs do not require fully precise estimates of
ancestral recombination events,
and allow us to directly express important forms of temporal uncertainty.

@@

% This is all a bit boring, maybe, but I think the people we want to
% influence into using tskit will need to have this spelled out for them.
Rigorously defining interchange formats~\cite[e.g.][]{kelleher2019htsget}
is difficult and time-consuming,
and no matter how precise the specification, in practise
it is the \emph{implementations} that determine how well methods
interoperate. The BAM read alignment format~\citep{li2009sequence}
is an instructive example. Originally developed as part of the
1000 Genomes project~\citep{10002015global} to address the fragmented
software ecosystem that existed at the time~\citep{danecek2021twelve},
BAM has since become ubiquitous in bioinformatics pipelines.
The excellent interoperability between methods
exchanging alignment data is largely attributable to the
success of \texttt{htslib}~\citep{bonfield2021htslib}, the
software library that \emph{implements} BAM and several other
foundational bioinformatics file formats.
Today, there are thousands of software projects
using \texttt{htslib}~\citep{bonfield2021htslib},
and it this shared use of community software
infrastructure that guarantees the smooth flow of data between
applications.
The emerging ARG software ecosystem could similarly benefit
from the adoption of such shared community infrastructure
to handle the mundane and time-consuming details of data interchange.
The \texttt{tskit} library (Section~\ref{sec-efficiency})
is a high-quality open-source gARG implementation,
with proven efficiency and
scalability~\citep[e.g.][]{anderson2022genes,zhan2023towards},
that is already in widespread use.
Adopting it as a community standard may ease software implementation burden
on researchers, freeing their time to address the many fascinating open
questions and challenges that exist.

@@

\section*{Data Availability}
The public GitHub repository at
\url{https://github.com/tskit-dev/what-is-an-arg-paper} can be
used to reproduce all figures and tables in this paper.
In particular this includes
the ARG used in Fig.~\ref{fig-ancestry-resolution};
the simulation code and functions
used to generate Fig.~\ref{fig-simplification};
and for Fig.~\ref{fig-inferred-args},
the software versions, parameter settings, and
(where necessary) functions to convert
software outputs to the \texttt{tskit} gARG format.

@@

% Note: good to keep to paras here to logically separate the two important
% points.
The second property of how evolutionary trees
are conventionally represented that is unhelpful in the context of ARGs is their
focus on branching points (coalescences), i.e.\ nodes that have two or more children.
As the introductory paragraph of this section emphasised,
parent-child relationships are what fundamentally define a tree,
and branching points can be seen as incidental. This is reflected
by the oriented tree encoding which simply stores the local
parent-child relationships, and does not, for example,
directly tell us how many children a particular node has.
The local tree at a given position records the \emph{path} through
the ARG; if this path omits nodes that are not
branching points (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
we lose information about the ARG.
We return to this point in the following two sections,
when we discuss ``locally unary'' nodes and the simplification process.

>>>>

We show that a synthesis of these views can be achieved by thinking
in terms of genomes and transmission
intervals.
% Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
% based on exhaustively detailing the ``events'' that occur
% in the history of a sample,
%are insufficient to describe the outputs of current methods.
We argue that classical ARG formulations, which assume
that all events can be known and precisely estimated,
are fundamentally unsuited to the realities
of modern, population-scale datasets.
We describe an alternative formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

The breakthroughs in scale achieved by recent
methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history.
% This is a digression from the maing point of the paragraph, so
% making it a parenthetical aside.
It is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

We highlight this problem and show how to address it by a simple formal definition of an
ARG data structure, based on recording the intervals of
genetic inheritance between specific genomes.
We call this the ``genome ARG'', or gARG encoding. We
contrast this with the classical formal definition of an ARG, based on
recording common ancestor and recombination events, which we refer to as the
``event ARG'' or eARG encoding. We show that the new gARG encoding is a
substantial generalisation of the classical eARG approach,  providing much more
flexibility in how genetic inheritance can be represented, and encompasses the
outputs of modern methods. We show that the gARG approach can represent many
different types of approximation, in particular allowing us to systematically
describe uncertainty about the temporal ordering of multiple recombinations.

@@

% % I feel like sophisticated readers would at this point be saying,
% % ah, yeah, that's basically what an ARG is just in different words;
% % what's the point of this paper?
The genome ARG framework defined here is
in many ways simply a clarification of existing treatments
\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
adding concrete details to describe the
differential inheritance of genetic material between genomes.
% % This is definitely needed for the Nicks, who are always thinking
% % about the deeper processes and are quite happy being loosey-goosey
% % with representation details
In the next section we contrast this gARG representation with a classical event-based (eARG) representation of the same underlying process. The choice between using a gARG or an eARG to represent the general process of ancestral inheritance has 
% Again, need to remind the reader that there's an actual point
% to all this getting stuck into the details.
important
consequences not just when addressing the practical need to store and compare ARGs, but more fundamentally when establishing goals for inferring ARGs and evaluating the success of such inferences.

\section{Event ARGs}
\label{sec-eARG}
A common, classical view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability},
interprets nodes not as genomes but as historical \emph{events}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions).
This eARG encoding is the basis of the output formats created by multiple ARG inference tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this case, two types of internal node exist in the graph. These
represent the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

% eARGs are representationally for limited interchange
While this approach of annotating recombination nodes with a
breakpoint in an eARG is a concise and elegant way of describing realisations
of the coalescent, it is inherently limited when
implemented literally. The eARG encoding explicitly models only
two different types of event; thus anything that is not a single crossover
recombination or common ancestor event must be incorporated
either in a roundabout way using these
events, or by adding new types of event to the encoding. For example, gene
conversion could be accommodated either by stipulating a third type of event
(annotated by two breakpoints and corresponding traversal conventions for
recovering the local trees) or by two recombination nodes joined by a
zero-length edge. The gARG encoding described in the previous
section offers a simpler and more direct solution.

@@

\section{Ancestral material and sample resolution}
\label{sec-ancestral-material}
Ancestral material~\citep{wiuf1999ancestry,wiuf1999recombination}
is a key concept in understanding the overall inheritance structure
of an ARG (here, and throughout, we use the general term ``ARG'' when
the details of the specific encoding are not important).
It denotes the genomic intervals ancestral to a set of samples
on the edges of an ARG.
For example, in Fig.~\ref{fig-arg-in-pedigree} we have
four sample genomes, \noderef{a}--\noderef{d}. As we
trace their genetic ancestry into the previous generation
(\noderef{e}--\noderef{h}), we can think of their ancestral
material propagating through the graph
backwards in time. In the region $[2, 7)$, there is a
local coalescence where nodes \noderef{a} and \noderef{c}
find a common ancestor in \noderef{f}. Thus, in this region,
we have three genome segments that are ancestral to the
four samples. Fig.~\ref{fig-arg-in-pedigree}A
illustrates this by (shaded) ancestral material being present
in only three nodes (\noderef{f}, \noderef{g}, and \noderef{h}) in this region,
while node \noderef{e} is blank
as it carries \emph{non-ancestral} material.
This process of local coalescence continues through the
graph, until all samples reach their most recent common
ancestor in node \noderef{n}.

@@

Efficiently propagating and resolving ancestral material for
a sample through a pre-existing graph is a well-studied problem,
and central to recent advances in individual-based forward-time
simulations~\citep{kelleher2018efficient,haller2018tree}.
In contrast to the usual ``retrospective'' view of ARGs
discussed so far, these methods record an ARG forwards in
time in a ``prospective'' manner. Genetic inheritance relationships
and mutations are recorded exhaustively, generation-by-generation,
leading to a rapid build-up of information, much of which
will not be relevant to the genetic ancestry of a future population.
This redundancy is periodically removed using the ``simplify''
algorithm~\citep{kelleher2018efficient}, which propagates and
resolves ancestral material.
Efficient simplification is the key enabling factor for
this prospective-ARG based approach to forward-time simulation,
which can be orders of magnitude faster than standard
sequence-based methods
(see Section~\ref{sec-ARG-simplification} for
other applications of ARG simplification).
% Define "sample-resolved" here for later use
We refer to a gARG that has been simplified with respect to a set of
samples, such that the inheritance annotations on its edges contain
no non-ancestral material,  as sample-resolved.

@@

Any eARG can be converted to a sample-resolved gARG
via a two-step process illustrated in Fig.~\ref{fig-ancestry-resolution}.
The first step is to take the input eARG (Fig.~\ref{fig-ancestry-resolution}A),
duplicate its graph topology, and then add inheritance annotations
to each of the gARG's edges (Fig.~\ref{fig-ancestry-resolution}B) as follows.
If a given node is a common ancestor event, we annotate the single
outbound edge with the interval $[0,L)$, for a genome of length $L$. If the
node is a recombination event with a breakpoint $x$, we annotate the two
outbound edges respectively with the intervals $[0, x)$ and $[x, L)$. These
inheritance interval annotations are clearly in one-to-one correspondence with
the information in the input eARG. They are also analogous to the
inheritance intervals we get on the edges in a prospective gARG
produced by a forward-time simulation, which are concerned with recording
the direct genetic relationship between a parent and child genome and are not
necessarily minimal in terms of the ancestral material of a sample.
Thus, the final step is to use the ``simplify'' algorithm to resolve the
ancestral material of the samples (Fig.~\ref{fig-ancestry-resolution}C).

@@

There are a number of tools for inferring genetic genealogies,
and the graph structures that they produce differ in important aspects.
These varying representations have all been termed ``ARGs'' by different authors,
and each describes more than a simple sequence of uncorrelated trees.
By way of illustration, Fig.~\ref{fig-inferred-args} depicts the output of four
different ``ARG inference'' tools run on a common dataset,
with the resulting ARGs converted into a standard gARG format (cite tskit).

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate, converted to the standard \emph{tskit} gARG encoding. See
Appendix \ref{sec-example-inferred-args} for details of these methods.
Edge colours indicate time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Bottom row graphics show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines denote breakpoints between
local trees.
}
\end{figure}

As is well-known (cite, perhaps Gusfield?), many different ancestral graph topologies are compatible with the same resulting
sequence data. Hence it is unsurprising that the four graphs differ,
for example in the numbers of nodes and their relationship to each other.
However, more detailed inspection of Fig.~\ref{fig-inferred-args} reveals a
diversity of representation that is only possible via the gARG encoding.
For instance, although the most parsimonious ARG in Fig.~\ref{fig-inferred-args}A
has far fewer ancestral events than an ARG instance sampled
from from the coalescent with recombination in Fig.~\ref{fig-inferred-args}B,
both consist only of nodes with either two parents or two children (but not both).
These nodes correspond to recombination and common ancestor events in the equivalent eARG.
In contrast, more scalable heuristic inference methods (Fig.~\ref{fig-inferred-args}C and D)
create ARGs in which a node can simultaneously have more that than two children and more than two parents.
Note also that the first three methods infer nodes that have a
``coalescence span'' of less than 100\%, indicating that in the local trees, ancestral lineages exist 
that pass through these nodes without the nodes marking a coalescent point. Such ``locally-unary''
nodes are an notable feature of the local trees defined by an ARG
(see Appendix \ref{sec-locally-unary-edges}).

These differences, in particular the number of parents and children, and the coalescence span, reflect the
degree of precision represented in the inferred ARGs. Indeed, it is important to note that the gARG encoding
allows a single underlying genetic genealogy to be represented at
differing levels of detail. The concept is illustrated in Fig.~\ref{fig-simplification},
which shows different levels of ``simplification'' (Appendix~\ref{sec-ARG-simplification}) of the
same underlying simulated ARG. 
The full ARG, with all coalescent and recombination events represented by
separate genomes, is shown in Fig.~\ref{fig-simplification}A. Simpler representations can be formed by removing ``unknowable'' nodes (Fig.~\ref{fig-simplification}B) and 
collapsing multiple recombinations into a single child or multiple coalescences into
a single parent (Fig.~\ref{fig-simplification}C, also see Appendix~\ref{sec-cell-lineages-and-args}).
Finally, if we chose to represent only local coalescence points (Fig.~\ref{fig-simplification}D),
we end up with a ``fully simplified'' ARG
of the sort inferred in Fig.~\ref{fig-inferred-args}D. Note that in all cases, the equivalent
trees only differ in the locally-unary nodes they contain; topological relationships
between the samples remain identical.

@@

The ability to represent an ARG to differing degrees of precision is a powerful
feature of gARGs. In particular, when inferring ARGs from genome
sequencing data, the timing, positions, and even the number of recombination
events is generally not possible to infer precisely. For example, under
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation and recombination
rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
position can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region. Because a gARG
can encode a diversity of ARG structures, it
allows this fundamental uncertainty in inference to be appropriately represented
(Appendix~\ref{sec-precision}.
% For general
% ARGs, for instance those representing pedigrees, or genealogies of organisms with
% low mutation rates and complex patterns of recombination, such inference can be
% even more challenging.
%The fact that the eARG encoding \emph{requires}
%precise information about recombination is therefore a fundamental limitation.

\section{Discussion}
\label{sec-discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in inference methods, their evaluation, and their application to biological questions.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As reviewed in Appendix~\ref{sec-arg-history}, the term
``ancestral recombination graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definition
within the context of stochastic
processes and instead use the term as a more general representation of any
recombinant genetic ancestry seems useful.
Our intent here is to clarify and systematise this decoupling. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG (gARG) encoding
is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG) encoding, which is more limited in what can be described.
Importantly, gARGs do not require fully precise estimates of
ancestral recombination events,
and allow us to directly express important forms of temporal uncertainty.

@@

% This is all a bit boring, maybe, but I think the people we want to
% influence into using tskit will need to have this spelled out for them.
Rigorously defining interchange formats~\cite[e.g.][]{kelleher2019htsget}
is difficult and time-consuming,
and no matter how precise the specification, in practise
it is the \emph{implementations} that determine how well methods
interoperate. The BAM read alignment format~\citep{li2009sequence}
is an instructive example. Originally developed as part of the
1000 Genomes project~\citep{10002015global} to address the fragmented
software ecosystem that existed at the time~\citep{danecek2021twelve},
BAM has since become ubiquitous in bioinformatics pipelines.
The excellent interoperability between methods
exchanging alignment data is largely attributable to the
success of \texttt{htslib}~\citep{bonfield2021htslib}, the
software library that \emph{implements} BAM and several other
foundational bioinformatics file formats.
Today, there are thousands of software projects
using \texttt{htslib}~\citep{bonfield2021htslib},
and it is this shared use of community software
infrastructure that guarantees the smooth flow of data between
applications.
The emerging ARG software ecosystem could similarly benefit
from the adoption of such shared community infrastructure
to handle the mundane and time-consuming details of data interchange.
The \texttt{tskit} library (Section~\ref{sec-efficiency})
is a high-quality open-source gARG implementation,
with proven efficiency and
scalability~\citep[e.g.][]{anderson2022genes,zhan2023towards},
that is already in widespread use.
Adopting it as a community standard may ease software implementation burden
on researchers, freeing their time to address the many fascinating open
questions and challenges that exist.

@@

\section*{Data Availability}
The public GitHub repository at
\url{https://github.com/tskit-dev/what-is-an-arg-paper} can be
used to reproduce all figures and tables in this paper.
In particular this includes
the ARG used in Fig.~\ref{fig-ancestry-resolution} as well as
the simulation code and functions
used to generate Fig.~\ref{fig-inferred-args} and Fig.~\ref{fig-simplification}.
This includes the software versions, parameter settings, and
(where necessary) functions to convert
software outputs to the \texttt{tskit} gARG format.

@@

% Note: good to keep to paras here to logically separate the two important
% points.
The second property of how evolutionary trees
are conventionally represented that is unhelpful in the context of ARGs is their
focus on branching points (coalescences), i.e.\ nodes that have two or more children.
As the introductory paragraph of this section emphasised,
parent-child relationships are what fundamentally define a tree,
and branching points can be seen as incidental. This is reflected
by the oriented tree encoding which simply stores the local
parent-child relationships, and does not, for example,
directly tell us how many children a particular node has.
The local tree at a given position records the \emph{path} through
the ARG; if this path omits nodes that are not
branching points (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
information about the ARG is lost.
We expand on this point in Appendices~\ref{sec-locally-unary-edges} and
\ref{sec-ARG-simplification}
where we which discuss ``locally unary'' nodes and the simplification process.
